---
title: "NMRcycling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NMRcycling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(AlpsNMR)
library(factoextra)
library(readxl)
library(prospectr)
library(httpuv)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(IMIFA)
library(gridExtra)
library(tidyr)
library(NMRcycling)
```

# Introduction

# Data analysis

## Read data

### Read NMR data

```{r get_dataset_names}
# get all folders
datasetnames <- list.dirs(path = system.file("extdata", 
                                             package = "NMRcycling"),
                          recursive = FALSE)

# get the sample name from the folder name
sample_names <- basename(datasetnames)
```

```{r read_data, message=FALSE, warning=FALSE}
# read all data
datasets <- nmr_read_samples(sample_names = file.path(datasetnames, "10"))

# the experiment names are incorrect, fix this here
datasets$metadata <- lapply(datasets$metadata, function(x) {
  x$NMRExperiment <- sample_names
  return(x)
})
```

```{r show_sample, fig.width=7, fig.height=5}
# interpolate 1D NMR spectra
datasets <- nmr_interpolate_1D(samples = datasets, 
                               axis = c(min = -0.5, max = 10, by = 2.3E-4))

# show raw NMR plots
plot(datasets, 
     NMRExperiment = c("Sample9999")) +
  theme_minimal()
```

### Read meta data

```{r metadata}
# read the meta data, meta data is in hte package NMRcycling
meta <- read_excel(path = system.file("metadata", "metadata.xlsx",
                                      package = "NMRcycling"))

# change some columns to factor
meta$Timepoint <- factor(meta$Timepoint)

# set names for adding meta data
datasets$metadata$external$path <- sample_names

# add meta data to the dataset
datasets <- nmr_meta_add(nmr_data = datasets, 
                         metadata = meta, 
                         by = c("path" = "Path"))
```

## Pre-processing

Remove regions from the spectra. 

* water peak region between 4.5 and 5.0 ppm
* points outside of the range where we expect metabolite peaks, downfield of 10 ppm
and upfield of -0.5 ppm.

```{r exclude_regions}
# define regions to exclude
regions <- list(water = c(4.5, 5.0), 
                upfield = c(-10, 0.5), 
                downfield = c(10, 20))

# exclude the regions
datasets <- nmr_exclude_region(samples = datasets, 
                               exclude = regions)
```

## Pre-treatment

### Normalization

```{r normalization, fig.width=7, fig.height=5}
# normalization
datasets <- nmr_normalize(samples = datasets, 
                          method = c("pqn"))

# show normalized spectra
plot(datasets, 
     NMRExperiment = "Sample9999") +
  theme_minimal()
```

### HMDB library

The `AlpsNMR` package contains the chemical shift database from the Human Metabolome Database (HMDB). You can load 
database with the following command.

```{r load_hmdb}
load(file = system.file("data", "hmdb.rda",
                        package = "AlpsNMR"))
```

On the website of the HMDB we can find the indices (accession numbers) of the ketone bodies: 11, 60 and 1659 for 3-hydroxybutyrate, acetoacetate and acetone respectively.

```{r show_chemical_shift}
hmdb$pos_in_ppm <- hmdb$pos_in_ppm - 1.47 + 1.486
hmdbket <- hmdb[grepl("HMDB0000011|HMDB0000060|HMDB0001659", hmdb$Accession), ]
hmdbket[, c("Metabolite", "pos_in_ppm", "Accession", "Urine")]
```

```{r extract_metabolites}
# select only the Exercise days
dsexcday <- datasets[datasets$meta$external$Day == "Exercise day"]
# select the 3 shifts for the 3 metabolites
hmdbsel <- hmdbket[c(1, 5, 7), ]

# set the ranges for plotting
rn1 <- hmdbsel[[1, "pos_in_ppm"]] + c(-1, 1) * 0.03
rn2 <- hmdbsel[[2, "pos_in_ppm"]] + c(-1, 1) * 0.03
rn3 <- hmdbsel[[3, "pos_in_ppm"]] + c(-1, 1) * 0.03
```

```{r show_metabolites, fig.width=7, fig.height=5}
pl1 <- plot(dsexcday, 
            chemshift_range = rn1, 
            color = "Timepoint") + 
  ggtitle(hmdbsel[[1, "Metabolite"]])
pl2 <- plot(dsexcday, 
            chemshift_range = rn2, 
            color = "Timepoint") + 
  ggtitle(hmdbsel[[2,"Metabolite"]])
pl3 <- plot(dsexcday, 
            chemshift_range = rn3, 
            color = "Timepoint") + 
  ggtitle(hmdbsel[[3,"Metabolite"]])

pl1 + pl2 + pl3 +
  plot_layout(guides = "collect") &
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "bottom")
```

### Binning

```{r binning, fig.width=7, fig.height=5}
# make sure to keep the original dataset
ds <- datasets
# bin the data
ds$data_1r <- binning(X = ds$data_1r, 
                      bin.size = 200)

# bin the axis
ds$axis <- binning(X = ds$axis, 
                   bin.size = 200)

# show binned spectra
plot(ds, 
     NMRExperiment = "Sample9999") +
  theme_minimal()
```

## Analysis

### t-test

Do a paired t-test on all bins.

```{r}
# make data.frame
df_data <- cbind.data.frame(ds$metadata$external, ds$data_1r)

# remove early timepoints, i.e. timepoint 0 and 1 and remove the pooled sample
remove_pool <- df_data$Day != "Pool"
remove_early <- !(df_data$Timepoint == "0" | 
                    df_data$Timepoint == "1")
df_data <- df_data[remove_pool & remove_early, ]

# sort on timepoint to make sure that below the subsettings creates the correct pairs
df_data <- df_data[order(df_data$Timepoint), ]

# calculate the p-value and log2 fold change
df_res <- data.frame(
  bin = round(ds$axis, 2),
  logp = -log10(sapply(df_data[, -c(1:6)], function(x) t.test(x = x[df_data$Day == "Exercise day"],
                                                              y = x[df_data$Day == "Rest day"],
                                                              paired = TRUE)$p.value)),
  logfc = log2(colMeans(df_data[df_data$Day == "Exercise day", -c(1:6)]) / colMeans(df_data[df_data$Day == "Rest day", -c(1:6)]))
)
```

```{r volcano plot, fig.width=7, fig.height=5}
ggplot(data = df_res,
       aes(x = logfc,
           y = logp,
           label = bin)) +
  geom_point() +
  geom_text_repel() +
  geom_vline(xintercept = c(-1, 1), 
             linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05 / nrow(df_res)), 
             linetype = "dashed") +
  theme_minimal()
```

### PCA analysis

```{r PCA}
# get group information
groups <- ds$metadata$external$Phase

# create data.frame for PCA analysis
df <- ds$data_1r
rownames(df) <- ds$metadata$external$RowID
colnames(df) <- round(ds$axis, 2)

# Remove pooled sample
df2 <- df[rownames(df) != "Pool",]
groups2 <- groups[groups != "Pool"]

# Mean centering and Pareto scaling 
df2 <- apply(X = df2,
             MARGIN  = 2,
             function(x) { (x - mean(x)) / sqrt(sd(x)) })

df2 <- df2 * 500.0

# PCA
res <- (prcomp(df2))
```

Show the scores plot.

```{r pca_scoresplot, fig.width=7, fig.height=5}
# show scores plot
fviz_pca_ind(X = res,
             axes = c(1, 2),
             col.ind = groups2, # color by groups
             #palette = c("#00AFBB",  "#FC4E07", "#00AFBB",  "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups",
             repel = TRUE
)
```

Show the biplot.

```{r pca_biplot, fig.width=7, fig.height=5}
fviz_pca_biplot(X = res,
                repel = TRUE,
                axes = c(1, 2),
                col.var = "contrib", # Variables color
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                select.var = list(contrib = 10)
)
```

#### Interesting variables

Select bin **3.19** and have a closer look.

```{r bin319, fig.width=7, fig.height=5}
# should rename df
sel <- which(colnames(df) == "3.19") + c(-1, 0, 1)
rn <- ds$axis[sel]
rn <- 0.5 * c(rn[1] + rn[2], rn[2] + rn[3])

plot(datasets[datasets$meta$external$Day == "Exercise day"], 
     chemshift_range = rn, 
     color = "Timepoint") +
  theme_minimal()
```

Look up the chemical shift in HMDB.

```{r}
hmdbsel <- hmdb[hmdb$pos_in_ppm > min(rn) &
                  hmdb$pos_in_ppm < max(rn), ]
hmdbsel <- hmdbsel[is.na(hmdbsel$J_constant), ]
hmdbsel <- hmdbsel[grepl(pattern = "HMDB000....", 
                         x = hmdbsel$Accession), ]

# show the result
hmdbsel[, c("Metabolite", "pos_in_ppm", "Accession", "Urine")]
```

Focus on **L-Acetylcarnitine**.

```{r}
hmdb[hmdb$Accession == "HMDB0000201", c("Metabolite", "pos_in_ppm", "Accession", "Urine")]
```

Check the spectra.

```{r acetylcarnitine, fig.width=7, fig.height=5}
hmdbsel <- hmdb[hmdb$Accession == "HMDB0000201", ]
hmdbsel <- hmdbsel[round(hmdbsel$pos_in_ppm * 1000) != 3196, ]
# extract the chemical shift
pos <- hmdbsel[[1, "pos_in_ppm"]]
rn <- 0.02 * c(-1, 1) + pos

plot(datasets[datasets$meta$external$Day == "Exercise day"], 
     chemshift_range = rn, 
     color = "Timepoint") +
  theme_minimal()
```

Select bin **1.3** and have a closer look.

```{r bin12, fig.width=7, fig.height=5}
# should rename df
sel <- which(colnames(df) == "1.3") + c(-1, 0, 1)
rn <- ds$axis[sel]
rn <- 0.5 * c(rn[1] + rn[2], rn[2] + rn[3])

plot(datasets[datasets$meta$external$Day == "Exercise day"], 
     chemshift_range = rn, 
     color = "Timepoint") +
  theme_minimal()
```

Look up the chemical shift in HMDB.

```{r}
hmdbsel <- hmdb[hmdb$pos_in_ppm > min(rn) &
                  hmdb$pos_in_ppm < max(rn), ]
# hmdbsel <- hmdbsel[is.na(hmdbsel$J_constant), ]
hmdbsel <- hmdbsel[hmdbsel$Urine == TRUE,]
hmdbsel <- hmdbsel[grepl(pattern = "HMDB000....", 
                         x = hmdbsel$Accession), ]

# show the result
hmdbsel[, c("Metabolite", "pos_in_ppm", "Accession", "Urine")]
```

Have a closer look at **Capric acid**

```{r}
hmdb[hmdb$Accession == "HMDB0000511", c("Metabolite", "pos_in_ppm", "Accession", "Urine")]
```

Check the spectra for the 0.866 ppm.

```{r capric_acid, fig.width=7, fig.height=5}
hmdbsel <- hmdb[hmdb$Accession == "HMDB0000511",]
# select the smallest chemical shift
pos <- min(hmdbsel[, "pos_in_ppm"])
rn <- 0.02 * c(-1, 1) + pos

plot(datasets[datasets$meta$external$Day == "Exercise day"], 
     chemshift_range = rn, 
     color = "Timepoint") +
  theme_minimal()
```

## Peak fitting

### Creatinine

```{r fig.width=7, fig.height=5}
sp <- datasets[datasets$metadata$external$RowID == "Pool"]$data_1r
ax <- datasets[datasets$metadata$external$RowID == "Pool"]$axis
sel <- (ax > 3.03) & (ax < 3.07)

pl <- peakfit(ax[sel], 3.051, 1 / 600, 0, 1, 0, sp[sel])

spfit <- do.call("peak", pl)
dffit <- data.frame(ax = ax[sel], 
                    sp = sp[sel], 
                    spfit)

ggplot(data = dffit,
       aes(x = ax,
           y = sp)) + 
  geom_point() + 
  geom_line(aes(y = spfit)) +
  theme_minimal()
```

```{r}
creatinine <- NULL
id <- datasets$metadata$external$RowID
for(t in id) {
  pos <- which(datasets$metadata$external$RowID == t)
  dffit <- data.frame(sp = datasets[pos]$data_1r[sel], 
                      ax = datasets[pos]$axis[sel])
  
  result <- peakfitsimple(dffit$ax, pl$v0, pl$fwhh, pl$p, pl$a, pl$b, dffit$sp)
  
  result$v <- NULL
  result$id <- t
  creatinine <- rbind(creatinine, data.frame(result))
}
creatinine <- (creatinine$a / 3)
```

### beta-hydroxybutyrate

```{r fig.width=7, fig.height=5}
sp <- datasets[datasets$metadata$external$RowID == "ExerciseLate 7"]$data_1r
ax <- datasets[datasets$metadata$external$RowID == "ExerciseLate 7"]$axis
sel <- (ax > 1.208) & (ax < 1.22)

pl <- peakfit(ax[sel], 1.2102, 1 / 600, 0, 1, 0, sp[sel])

spfit <- do.call("peak", pl)
dffit <- data.frame(ax = ax[sel], 
                    sp = sp[sel], 
                    spfit)

ggplot(data = dffit, 
       aes(x = ax,
           y = sp)) + 
  geom_point() + 
  geom_line(aes(y = spfit)) +
  theme_minimal()
```

```{r}
bhb <- NULL
id <- datasets$metadata$external$RowID
for(t in id) {
  pos <- which(datasets$metadata$external$RowID == t)
  dffit <- data.frame(sp = datasets[pos]$data_1r[sel], 
                      ax = datasets[pos]$axis[sel])
  
  result <- peakfitsimple(dffit$ax, pl$v0, pl$fwhh, pl$p, pl$a, pl$b, dffit$sp)
  result$v <- NULL
  result$id <- t
  
  bhb <- rbind(bhb, data.frame(result))
}

bhb <- (2 * bhb$a / 3) / creatinine # number of protons, only half of doublet fitted
```

```{r fig.width=7, fig.height=5}
mets <- cbind(datasets$metadata$external, data.frame(bhb))
mets <- mets[mets$RowID != "Pool",]

ggplot(data = mets, 
       aes(x = Timepoint, 
           y = bhb, 
           group = Day, 
           color = Day)) + 
  geom_line(size = 2) +
  theme_minimal()
```

### Acetone

```{r fig.width=7, fig.height=5}
sp <- datasets[datasets$metadata$external$RowID == "ExerciseLate 7"]$data_1r
ax <- datasets[datasets$metadata$external$RowID == "ExerciseLate 7"]$axis
sel <- (ax > 2.222) & (ax < 2.25)

pl <- peakfit(ax[sel], 2.238, 1 / 600, 0, 1, 0, sp[sel])

spfit <- do.call("peak", pl)
dffit <- data.frame(ax = ax[sel], 
                    sp = sp[sel], 
                    spfit)

ggplot(data = dffit, 
       aes(x = ax,
           y = sp)) + 
  geom_point() + 
  geom_line(aes(y = spfit)) +
  theme_minimal()
```

```{r}
acetone <- NULL
id <- datasets$metadata$external$RowID
for(t in id) {
  pos <- which(datasets$metadata$external$RowID == t)
  dffit <- data.frame(sp = datasets[pos]$data_1r[sel], 
                      ax = datasets[pos]$axis[sel])
  
  result <- peakfitsimple(dffit$ax, pl$v0, pl$fwhh, pl$p, pl$a, pl$b, dffit$sp)
  result$v <- NULL
  result$id <- t
  
  acetone <- rbind(acetone, data.frame(result))
}

acetone <- ((acetone$a / 6) / creatinine) # number of protons
```

```{r fig.width=7, fig.height=5}
mets <- cbind(datasets$metadata$external, data.frame(acetone))
mets <- mets[mets$RowID != "Pool",]

ggplot(data = mets, 
       aes(x = Timepoint, 
           y = acetone, 
           group = Day, 
           color = Day)) + 
  geom_line(size = 2) +
  theme_minimal()
```

### Acetylcarnitine

```{r fig.width=7, fig.height=5}
sp <- datasets[datasets$metadata$external$RowID == "ExerciseLate 7"]$data_1r
ax <- datasets[datasets$metadata$external$RowID == "ExerciseLate 7"]$axis
sel <- (ax > 3.175) & (ax < 3.215)

pl <- peakfit(ax[sel], 3.198, 1 / 600, 0, 1, 0, sp[sel])

spfit <- do.call("peak", pl)
dffit <- data.frame(ax = ax[sel], 
                    sp = sp[sel], 
                    spfit)

ggplot(data = dffit, 
       aes(x = ax,
           y = sp)) + 
  geom_point() + 
  geom_line(aes(y = spfit)) +
  theme_minimal()
```

```{r}
acetylcarnitine <- NULL
id <- datasets$metadata$external$RowID
for(t in id) {
  pos <- which(datasets$metadata$external$RowID == t)
  dffit <- data.frame(sp = datasets[pos]$data_1r[sel], 
                      ax = datasets[pos]$axis[sel])
  
  result <- peakfitsimple(dffit$ax, pl$v0, pl$fwhh, pl$p, pl$a, pl$b, dffit$sp)
  result$v <- NULL
  result$id <- t
  
  acetylcarnitine <- rbind(acetylcarnitine, data.frame(result))
}

acetylcarnitine <- (acetylcarnitine$a / 9)/creatinine
```

```{r fig.width=7, fig.height=5}
mets <- cbind(datasets$metadata$external, data.frame(acetylcarnitine))
mets <- mets[mets$RowID != "Pool",]

ggplot(data = mets, 
       aes(x = Timepoint, 
           y = acetylcarnitine, 
           group = Day, 
           color = Day)) + 
  geom_line(size = 2) +
  theme_minimal()
```

### Dicarboxylate

```{r fig.width=7, fig.height=5}
sp <- datasets[datasets$metadata$external$RowID == "ExerciseLate 7"]$data_1r
ax <- datasets[datasets$metadata$external$RowID == "ExerciseLate 7"]$axis
sel <- (ax > 1.50) & (ax < 1.59)

pl <- peakfit(ax[sel], 1.551, 1 / 600, 0, 1, 0, sp[sel])

spfit <- do.call("peak", pl)
dffit <- data.frame(ax = ax[sel], 
                    sp = sp[sel], 
                    spfit)

ggplot(data = dffit, 
       aes(x = ax,
           y = sp)) + 
  geom_point() + 
  geom_line(aes(y = spfit)) +
  theme_minimal()
```

```{r}
dicarboxylate <- NULL
id <- datasets$metadata$external$RowID
for(t in id) {
  pos <- which(datasets$metadata$external$RowID == t)
  dffit <- data.frame(sp = datasets[pos]$data_1r[sel], 
                      ax = datasets[pos]$axis[sel])
  
  result <- peakfitsimple(dffit$ax, pl$v0, pl$fwhh, pl$p, pl$a, pl$b, dffit$sp)
  result$v <- NULL
  result$id <- t
  
  dicarboxylate <- rbind(dicarboxylate, data.frame(result))
}

dicarboxylate <- (dicarboxylate$a / 4)/creatinine
```

```{r fig.width=7, fig.height=5}
mets <- cbind(datasets$metadata$external, data.frame(dicarboxylate))
mets <- mets[mets$RowID != "Pool",]

ggplot(data = mets, 
       aes(x = Timepoint, 
           y = dicarboxylate, 
           group = Day, 
           color = Day)) + 
  geom_line(size = 2) +
  theme_minimal()
```

### Combine all compounds

```{r}
# combine all data.frames
mets <- cbind(datasets$metadata$external, data.frame(acetone, bhb, acetylcarnitine, dicarboxylate))
# make long for easy plotting
mets_long <- pivot_longer(data = mets, 
                          cols = c("acetone", "bhb", "acetylcarnitine", "dicarboxylate"),
                          names_to = "compound")

# remove the pooled sample
mets_long <- mets_long[mets_long$Day != "Pool", ]
```

```{r fig.width=7, fig.height=5}
ggplot(data = mets_long, 
       aes(x = Timepoint, 
           y = value, 
           group = compound, 
           color = compound)) + 
  geom_line(size = 1) +
  facet_wrap(. ~ Day) +
  theme_minimal() +
  theme(strip.text.x = element_text(face = "bold"),
        legend.position = "bottom")
```

# Sessioninfo

```{r}
sessioninfo::session_info()
```

